Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-02 review_article.Rmd

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Transforming statistics to visualisations considers individual observations or observation points aggregated into communities and geographical units.
Visualisations may be used for hypothesis generating and is often the first step in exploratory spatial data analysis. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed.

@@

@@

@@

> The aim here is to see the whole, in as much detail as possible, at a glance.

<!-- What is a cartogram? -->
Cartograms provide an alternative visualisation method for statistical and geographical information.
A cartogram alters the map base with the intention of improving the presentation of a statistic.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
For a single variable of interest, each map area is changed to emphasise the distribution by representing the corresponding value, in comparison to the value of the other areas [@ACCAC].

<!--
Medical application: TAAM 1965
Actual shapes of counties were sacrificed to the extent necessary to retain as far as possible, common boundaries, relative locations, and the approximate contour of the State.

There is no unique solution to the topology
problem of preparing a population-by-area
cartogram for a given population and its jurisdictional
boundaries. Although the map areas
of the various regions are fixed by their populations, their shapes and relative locations are not completely determined.

Other characteristics of the population
may also serve as the base for adjustment; for
example, sex, race, parity, socio-economic level,
or immune status.
-->

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D.
The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just as photographs of the globe from space.
Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information.
Diagrams that do not specify a projection can be considered to have some unknown projection. <!-- ??? cite tobler.-->

@@

A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult [@AOCAFE].
The distortion of a cartogram prevents the density obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just, giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

>>>>

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Transforming statistics to visualisations considers individual observations or observation points aggregated into communities and geographical units.
Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed.

@@

<!-- Displaying of cancer data on choropleths -->

@@

<!--
A particular spatial structure includes
the individuals affected and how they are connected in
communities, as well as the dynamics of these communities
and their organization into larger units. @SAMGIS
-->

<!-- Measures reported in cancer atlases -->

@@

<!-- What is a cartogram? -->
A cartogram alters the map base with the intention of improving the presentation of the statistic of interest. For a single variable of interest, each map area is changed to emphasise the distribution by representing the corresponding value, in comparison to the value of the other areas [@ACCAC].

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just as photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information.
Diagrams that do not specify a projection can be considered to have some unknown projection. <!-- ??? cite tobler.-->

<!-- Denominators -->
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed beyond choosing a projection. 
The purposeful distortion of the map space is beneficial when a uniform density of the map base is desired.
When a map base is transformed according to population density, population becomes a uniformly distributed background for the statistic presented [@GOINO].

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].

@SE Researchers mapping spatial patterns of disease
for small areas are confronted with a number of
challenges, including cartographic design, the
modifiable area unit problem, and also the “small
number” problem, where some places have no,
or few, disease events and/or a low population at
risk of contracting the disease.
